An evaluation framework for trajectory inference methods

### 2017-11-28 analysis/paper/paper.Rmd

Dozens[b] of trajectory inference methods have been developed over the last years, and more are being published almost every month (`r ref("sfig", "methods_over_time", "a")`). Initially most trajectory inference was focused on linear trajectories, with one start and one end point. However, the prospect of branching trajectories was already described early [@trapnell_dynamics_2014], and about half of the current methods can handle one or more bifurcations in the trajectory (`r ref("sfig", "methods_over_time", "b")`). Some methods can already handle more complex, cyclical and branching behaviour, and it is expected that in the future methods will be able to model even more complex behavior, such as multiple dynamic processes happening at parallel in a single-cell or the integration of datasets from different patients [@tanay_scaling_2017; @cannoodt_computational_2016].

@@

We gathered a list of `r sum(method_df$is_ti == "Yup", na.rm=TRUE)` TI methods from literature, and selected a subset of `r sum(!is.na(method_df$qc_score))` for evaluation, primarily based on their free availability and the presence of a programming interface. We characterized all methods in four different ways: underlying algorithm(s), implementation quality, prior information and possible trajectory structures (`r ref("fig", "method_characteristics")`). The underlying algorithms can be divided into several subcomponents, and these components are frequently shared between different algorithms. Some components, such as a minimal spanning tree or principal curves for cellular ordering, or PCA or diffusion map for dimensionality reduction are shared by more than 5 methods sFIG?. While not directly related to the quality of the inferred trajectory, the quality of the implementation is an important first evaluation metric, because good unit testing makes sure the implementation is correct, good documentation makes it easier for potential users to apply the method on their data, and overall good code quality makes it possible for other developers to adapt the method and extend it further. We therefore looked at the implementation of each method, and assessed their implementation quality using a transparent scoring scheme sNOTE. After initial scoring we contacted all the authors, to allow them to improve their method before publishing. While most methods lack at several metrics (`r ref("fig", "method_characteristics", "a")`), recent methods tend to have higher code quality sFIG. However, some implementation details are lacking for the majority of the methods and should receive extra attention from developers: code availability, documentation and unit testing (`r ref("fig", "method_characteristics", "a")`).

@@

### Data sources
![evaluation_overview](../../analysis/figures/evaluation_overview.png)
`r ref("fig", "evaluation_overview", anchor=TRUE)` Overview of the evaluation

We use three types of datasets for our evaluation
### Score metrics and parameter optimisation

@@

# Discussion

# Conclusion

@@

>>>>

Dozens[b] of trajectory inference methods have been developed over the last years, and more are being published almost every month (`r ref("sfig", "methods_over_time", "a")`). Initially, most trajectory inference was focused on linear trajectories, with one start and one end point. However, the prospect of branching trajectories was already described early [@trapnell_dynamics_2014], and about half of the current methods can handle one or more bifurcations in the trajectory (`r ref("sfig", "methods_over_time", "b")`). Some methods can already handle more complex, cyclical and branching behaviour, and it is expected that in the future methods will be able to model even more complex behavior, such as multiple dynamic processes happening at parallel in a single-cell or the integration of datasets from different patients [@tanay_scaling_2017; @cannoodt_computational_2016].

@@

We gathered a list of `r sum(method_df$is_ti == "Yup", na.rm=TRUE)` TI methods from literature, and selected a subset of `r sum(!is.na(method_df$qc_score))` for evaluation, primarily based on their free availability and the presence of a programming interface. We characterized all methods in four different ways: underlying algorithm(s), implementation quality, prior information and possible trajectory structures (`r ref("fig", "method_characteristics")`). The underlying algorithms can be divided into several subcomponents, and these components are frequently shared between different algorithms. Some components, such as a minimal spanning tree or principal curves for cellular ordering, or PCA or diffusion map for dimensionality reduction are shared by more than 5 methods sFIG?. While not directly related to the quality of the inferred trajectory, the quality of the implementation is an important first evaluation metric, because good unit testing makes sure the implementation is correct, good documentation makes it easier for potential users to apply the method on their data, and overall good code quality makes it possible for other developers to adapt the method and extend it further. We therefore looked at the implementation of each method, and assessed their quality using a transparent scoring scheme sNOTE. After initial scoring we contacted all the authors, to allow them to improve their method before publishing. While most methods lack at several metrics (`r ref("fig", "method_characteristics", "a")`), recent methods tend to have higher code quality sFIG. However, some implementation details are lacking for the majority of the methods and should receive extra attention from developers: code availability, documentation and unit testing (`r ref("fig", "method_characteristics", "a")`).

@@

### Evaluation overview
![evaluation_overview](../../analysis/figures/evaluation_overview.png)
`r ref("fig", "evaluation_overview", anchor=TRUE)` Overview of the evaluation

Our evaluation was structured in three main phases. In the first phase, we used toy data as a positive control for each method, assessing whether it can detect the correct trajectories in very simplistic data. Next, we generated synthetic data from a realistic model of gene regulation, combined with a simulation of the technical variation induced by single-cell RNA sequencing. We used this synthetic data in a parameter optimization procedure, where we used cross-validation between datasets to avoid overfitting on particular data set structures. Finally, in the third phase, we used the optimized parameters to score the performance of each method on real data. 
### Score metrics and parameter optimisation

@@

# Discussion
# Conclusion

@@

## Trajectory inference methods
## Toy data generation
## Synthetic data generation
Our workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods [@marbach_wisdom_2012], but required several additions 
## Real data processing
## Normalization
## Evaluation metrics
## Parameter optimization
